Concurrent Programming with CXL

### 2020-06-29 book/paper.tex
Latest

\glsadd{concurrent execution}
\glsadd{shared variable}
\glsadd{sequential execution}
\glsadd{determinism}
\glsadd{atomicity}

Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}%
program is mostly \emph{deterministic}.
\index{determinism}%
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the
execution of the various processes are \emph{interleaved}.
Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}%
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}%
because there is no other activity interfering with their execution.
Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other processes
may update memory locations while a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
\index{test}%
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

\glsadd{model checking}

@@

Model checking is not a replacement for formal verification.
\index{formal verification}%
Formal verification proves that a program is correct.  Model checking only
verifies that a program is correct for some \emph{model}.  Think of
a model as a test program.  Because model checking tries every possible
execution, this means that the test program needs to be
simple or it will take longer than we may care to wait for or run
out of memory.
In particular, the model needs to have a relatively small number of
reachable states.

So if model checking does not prove a program correct, why is it
useful?  To answer that question consider, for example, a sorting
algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all \emph{corner cases},
\index{corner case}%
including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
Note, however, that it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an \textbf{if}
statement.)

\glsadd{invariant}

@@

The syntax and semantics of CXL is similar to that of Python.
Python is familiar to many programmers and is easy to learn and
use.  We will assume that the reader is familiar with the basics
of Python programming.  We also will assume that the reader
understand some basics of machine architecture and how programs
are executed.  For example, we assume that the reader is familiar
with the concepts of CPU, memory, register, stack, and machine
instructions.

@@

\begin{figure}
\begin{code}
\cxlsource{Up}
\end{code}
\caption{[Up.cxl] Incrementing twice in parallel.}
\label{fig:inc}
\end{figure}

\glsadd{process}

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}%
running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \textit{count} and
an array \textit{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \textit{self}.
It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
After that, it verifies that the value of \textit{count} equals 2.  If not,
it reports the actual value of \textit{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.

\begin{quote}
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \textit{count} will always end up being 2?
\item What are the possible values of \textit{count} at the time of the
\textbf{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying machine architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

\glsadd{machine instruction}

@@

\glsadd{race condition}

@@

\glsadd{context}

@@

\glsadd{stack machine}

@@

CXL also generates an HTML file that allows exploring more details
of the execution interactively.
Open the suggested HTML file and you should see something like
Figure~\ref{fig:inchtml1}.

@@

\glsadd{critical section}
\glsadd{thread safety}

@@

\glsadd{mutual exclusion}

@@

\glsadd{lock}

@@

\glsadd{state}
\glsadd{process variable}

@@

\glsadd{step}
\glsadd{trace}

@@

\glsadd{spinlock}

@@

\glsadd{interlock instruction}

@@

\glsadd{blocked process}

@@

\glsadd{reader/writer lock}

@@

\glsadd{busy waiting}

@@

\glsadd{semaphore}

@@

\glsadd{producer/consumer problem}

@@

\glsadd{conditional critical section}

@@

\glsadd{starvation}

@@

\glsadd{property}

@@

\glsadd{fairness}

@@

\glsadd{monitor}
\glsadd{condition variable}

@@

\glsadd{deadlock}

@@

\glsadd{actor model}

@@

\glsadd{non-blocking synchronization}

@@

\glsadd{linearizable}

@@

\glsadd{barrier synchronization}

>>>>

\glsadd{concurrent execution}%
\glsadd{shared variable}%
\glsadd{sequential execution}%
\glsadd{determinism}%
\glsadd{atomicity}%

Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}%
program is mostly \emph{deterministic}.
\index{determinism}%
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the
execution of the various processes are \emph{interleaved}.
Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}%
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically} (indivisibly)
\index{atomicity}%
because there is no other activity interfering with their execution.
Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other processes
may update memory locations while a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
\index{test}%
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

\glsadd{model checking}%

@@

Model checking is not a replacement for formal verification.
\index{formal verification}%
Formal verification proves that a program is correct.  Model checking only
verifies that a program is correct for some \emph{model}.  Think of
a model as a test program.
Because model checking tries every possible execution, the test
program needs to be simple. Otherwise it may take longer than we
care to wait for or run out of memory.
In particular, the model needs to have a relatively small number of
reachable states.

So if model checking does not prove a program correct, why is it
useful?
To answer that question, let us consider a sorting algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all \emph{corner cases},
\index{corner case}%
including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
Note, however, that it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an \textbf{if}
statement.)

\glsadd{invariant}%

@@

The syntax and semantics of CXL is similar to that of Python.
Python is familiar to many programmers and is easy to learn and
use.  We will assume that the reader is familiar with the basics
of Python programming.  We also will assume that the reader
understands some basics of machine architecture and how programs
are executed.  For example, we assume that the reader is familiar
with the concepts of CPU, memory, register, stack, and machine
instructions.

@@

\glsadd{process}%

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}%
running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \textit{count} and
an array \textit{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \textit{self}.
It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
After that, it verifies that the value of \textit{count} equals 2.  If not,
it reports the actual value of \textit{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.
Note that although the processes are \emph{spawned} one at a time,
they will execute concurrently.  It is, for example, quite possible
that \texttt{incrementer(1)} finishes before \texttt{incrementer(0)}
even gets going.

\begin{figure}[h]
\begin{code}
\cxlsource{Up}
\end{code}
\caption{[Up.cxl] Incrementing twice in parallel.}
\label{fig:inc}
\end{figure}

\begin{quote}
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \textit{count} will always end up being 2?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying machine architecture are atomic (indivisible)---in fact they are.)
\end{itemize}
\end{quote}

\glsadd{machine instruction}%

@@

\glsadd{race condition}%

@@

\glsadd{context}%

@@

\glsadd{stack machine}%

@@

CXL also generates an HTML file that allows exploring more details
of the execution interactively.
Open the suggested HTML file and you should see something like
Figure~\ref{fig:inchtml1}.
If you hover the mouse over a machine instruction, it provides a
brief explanation.

@@

\glsadd{critical section}%
\glsadd{thread safety}%

@@

\glsadd{mutual exclusion}%

@@

\glsadd{lock}%

@@

\glsadd{state}%
\glsadd{process variable}%

@@

\glsadd{step}%
\glsadd{trace}%

@@

\glsadd{spinlock}%

@@

\glsadd{interlock instruction}%

@@

\glsadd{blocked process}%

@@

\glsadd{reader/writer lock}%

@@

\glsadd{busy waiting}%

@@

\glsadd{semaphore}%

@@

\glsadd{producer/consumer problem}%

@@

\glsadd{conditional critical section}%

@@

\glsadd{starvation}%

@@

\glsadd{property}%

@@

\glsadd{fairness}%

@@

\glsadd{monitor}%
\glsadd{condition variable}%

@@

\glsadd{deadlock}%

@@

\glsadd{actor model}%

@@

\glsadd{non-blocking synchronization}%

@@

\glsadd{linearizable}%

@@

\glsadd{barrier synchronization}%
